---
title: "p8105_hw3_sd3730"
author: "Stacey Dai"
date: "2024-10-13"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

#### Load the ny_noaa dataset
```{r}
data("ny_noaa")
```

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include weather station id, date of observation,  (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

#### Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is `NA`, indicating missingness. Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r cleandata}
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa = 
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```

#### Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

Below is a two-panel plot showing the average max temperature in January and in July in each station across years.

```{r twopanel}
ny_noaa %>% 
  group_by(id, year, month) %>% 
  filter(month %in% c(1, 7)) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

The mean temperature in January is much lower than the mean temperature in July for all stations and across all years. All stations appear to follow similar trends of temperature peaks and valleys within a month across the years, i.e. when one station has a high monthly mean temperature for a given year, most other stations also have a high monthly mean temperature for that year. We do see one uncharacteristically cold station in July of 1987 or 1988, as well as a few other less drastic outliers.

#### Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

Below we show a two-panel plot including (i) a hex plot of `tmax` vs `tmin` for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values (in mm) greater than 0 and less than 100 separately by year. 

```{r ridgeplot}
hex = 
  ny_noaa %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa %>% 
  filter(snow < 100, snow > 0) %>%
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```

From the hex plot we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution. In relatively rare cases, it seems that `tmax` is less than `tmin`, which raises questions about data recording and quality.

From the ridge plot, we see a multimodal density of snowfall within a given year. Most stations see between 0 and  35 mm of snow in a year. Then there is a another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm. It is likely this multimodality stems from the conversion of measurements in one system (fractions of an inch) to another (using the metric system), which was also noted in the table of common values. 

# Problem 2

#### Load the dataset
```{r p2_setup, message = FALSE}
library(readr)
demographic_data = read_csv (file = "./data/nhanes_covar.csv", skip = 4)
names(demographic_data) = c("SEQN", "sex", "age", "BMI", "education")

accelerometer_data = read_csv (file = "./data/nhanes_accel.csv")
```

#### Clean and merge the datasets
We will filter out participants under 21 years of age and handle missing values. We will also tidy the data so that it is long-form and each observation represents a minute-by-minute MIMS reading for a participant.

```{r p2_cleanup}
library(tidyr)

demographic_data_clean = demographic_data %>%
  filter(age >= 21) %>%
  drop_na()

accelerometer_data_long = accelerometer_data %>%
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    values_to = "value"
  )
```

After cleaning both datasets, merge them on the participant ID to create a single dataset.
```{r p2_merge}
library(dplyr)

merged_data = accelerometer_data_long %>%
  left_join(demographic_data_clean, by = "SEQN") %>%
  mutate(
    sex = factor(sex, levels = c(1,2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), 
                       labels = c("Less than HS", "HS equivalent", "More than HS"))
  )
```

#### Table of Men and Women by Education Category

```{r p2_table}
library(knitr)

education_sex_table = merged_data %>%
  filter(!is.na(education)) %>%
  distinct(SEQN, .keep_all = TRUE) %>%
  count(education, sex) %>%
  pivot_wider(names_from = sex, values_from = n, values_fill = 0)

kable(education_sex_table, caption = "Number of Men and Women in Each Education Category")
```

The table shows that among participants with less than a high school education, there are slightly more females (28) than males (27). This suggests a relatively balanced distribution between genders in this category. 
In the HS equivalent category, the male count (35) is higher than the female count (23). This may indicate a trend where more males than females have completed high school, though the gap is not very large.
The category with the highest counts is the "More than HS" category, where there are 56 males compared to 59 females. This demonstrates that women in this group slightly outnumber men. The data suggests that a significant portion of both genders continues their education beyond high school, with a slight female majority in higher education.
Overall, the data indicates that females are represented slightly more in the less than HS and more than HS categories, while males are slightly more represented in the HS equivalent category. This could point to varying educational trends and social factors affecting education levels by gender within this population.
This distribution can prompt further analysis into factors that influence educational attainment, such as socioeconomic status, cultural expectations, and access to educational resources for each gender.

#### Visualization of Age Distributions by Sex and Education
```{r p2_visual}
library(ggplot2)

cleaned_data = merged_data %>%
  filter(!is.na(age) & !is.na(education))

ggplot(merged_data, aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge", binwidth = 5) +
  facet_wrap(~education) +
  labs(
    title = "Age Distribution by Education Level and Sex",
    x = "Age",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal()
```

The histogram displays the age distributions of men and women across different education levels, allowing for a comparative analysis between sexes within each education category. In the 'Less than HS' category, there are slightly more women than men, with ages primarily concentrated in the lower age brackets. Conversely, in the 'More than HS' category, men and women appear to have similar age distributions, but women are slightly more prevalent in the older age groups.The data indicates that younger men tend to have a higher representation in the 'HS equivalent' and 'More than HS' categories compared to women, who show a more balanced distribution across ages in the same education levels.The age distribution for both sexes in the 'More than HS' category shows a pronounced peak around the mid-30s, suggesting that this group may be more engaged in higher education or career advancement at that age. Overall, this visualization underscores the importance of educational attainment and its relationship with age demographics, providing valuable insights for understanding the socio-economic landscape among different genders.

#### Total Activity by Age and Education, Separated by Sex

Aggregate MIMS values to total activity for each participant
```{r agg_mim}
total_activity_data <- merged_data %>%
  group_by(SEQN, age, sex, education) %>%
  summarize(total_activity = sum(value, na.rm = TRUE))
```

Plot total activity vs age, separated by sex and education level
```{r activity_plot}
ggplot(total_activity_data, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess") +
  facet_wrap(~education) +
  labs(
    title = "Total Daily Activity by Age, Sex, and Education Level",
    x = "Age",
    y = "Total Activity (MIMS)"
  ) +
  theme_minimal()
```

The plot illustrates total daily activity (in MIMS) against age for different educational levels, separated by sex. Each panel represents a unique education level category: "Less than HS," "HS equivalent," "More than HS," and "NA." There is a noticeable decline in total daily activity with increasing age across all education levels. This suggests that as participants grow older, their physical activity decreases, which aligns with common findings in physical activity research. In the "Less than HS" and "More than HS" panels, men (represented by the dark purple line) tend to exhibit higher total daily activity levels than women (represented by the yellow line) throughout most age ranges. However, in the "HS equivalent" panel, the gap between male and female activity levels appears narrower, suggesting that educational attainment may influence the disparity in activity levels between sexes. The trend lines for both sexes show a consistent downward trajectory as age increases, indicating that both men and women become less active as they age. Notably, the younger age group (20-30 years) shows higher activity levels, particularly for those with "More than HS" education, while older age groups experience a significant decline. Overall, the findings suggest that education level plays a significant role in physical activity levels across age groups, with men generally being more active than women. These insights can inform future health interventions aimed at promoting physical activity, particularly targeting older adults and addressing gender disparities in activity levels.

#### 24-Hour Activity Time Course by Education Level and Sex
Summarize MIMS values over the course of a day by hour
```{r sum_mim}
activity_time_course = merged_data %>%
  group_by(minute, sex, education) %>%
  summarise(mean_activity = mean(value, na.rm = TRUE)) %>%
  ungroup()
```

This is to convert the minute variable to hours.
```{r min_to_hours}
activity_time_course = activity_time_course %>%
  mutate(hour = as.numeric(minute) / 60)
```


