---
title: "p8105_hw3_sd3730"
author: "Stacey Dai"
date: "2024-10-13"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

#### Load the ny_noaa dataset
```{r}
data("ny_noaa")
```

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include weather station id, date of observation,  (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is `NA`, indicating missingness. Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r cleandata}
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa = 
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```