---
title: "p8105_hw3_sd3730"
author: "Stacey Dai"
date: "2024-10-13"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

# Problem 1

#### Load the ny_noaa dataset
```{r}
data("ny_noaa")
```

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include weather station id, date of observation,  (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is `NA`, indicating missingness. Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r cleandata}
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa = 
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin))
```

Below is a two-panel plot showing the average max temperature in January and in July in each station across years.

```{r twopanel}
ny_noaa %>% 
  group_by(id, year, month) %>% 
  filter(month %in% c(1, 7)) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

The mean temperature in January is much lower than the mean temperature in July for all stations and across all years. All stations appear to follow similar trends of temperature peaks and valleys within a month across the years, i.e. when one station has a high monthly mean temperature for a given year, most other stations also have a high monthly mean temperature for that year. We do see one uncharacteristically cold station in July of 1987 or 1988, as well as a few other less drastic outliers.

Below we show a two-panel plot including (i) a hex plot of `tmax` vs `tmin` for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values (in mm) greater than 0 and less than 100 separately by year. 

```{r ridgeplot}
hex = 
  ny_noaa %>% 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_hex()

ridge = 
  ny_noaa %>% 
  filter(snow < 100, snow > 0) %>%
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

hex + ridge
```
